---
title: "Report"
author: "Lucas Steltenpohl"
date: "2023-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggfortify)
library(psych)
library(mice)
library(MASS)
library(pROC)
library(pls)
select = dplyr::select
```

# NFL Quarterback Predictive Analysis

This report is a statistical analysis of the performance and combine statistics of the National Football League (NFL) quarterbacks spanning from the seasons 2009-2016. The data set we utilized was compiled of multiple data sets of performance indicators that are critical in assessing a quarterback's effectiveness and impact on the field. Our data set encompasses a detailed examination of 103 quarterbacks and 22 variables, offering a unique insight into the changing dynamics and skill levels within the league during this period. The 22 variable names and descriptions are found in the table below.  

Variables:             |  Definitions:
---------------------- | ----------------------------
qb                     | Name of Quarterback
att                    | Number of passing attempts made throughout their career
cmp                    | Number of passing completions made throughout their career
yds                    | Number of total passing yards gained throughout their career
ypa                    | Average number of yards gained per attempt made 
td                     | Number of touchdowns thrown 
int                    | Number of interceptions thrown 
sack                   | Number of sacks taken 
loss                   | Number of yards lost when sacked
game_points            | Number of in game points responsible for
experience             | Number of years played in the NFL
Ht                     | Height of player in inches 
Wt                     | Weight of player in lbs
Forty                  | Player's combine 40 meter dash time in seconds 
Vertical               | Player's combine vertical jump in inches 
BroadJump              | Player's combine broad jump (horizontal) in inches
Cone                   | Player's combine 3 cone agility drill in seconds 
Shuttle                | Player's combine lateral speed test in seconds 
Year                   | Year in which the player was drafted 
Av                     | AV (Approximate Value) is a method created by pro-football-reference in an attempt to put a single number on the seasonal value of a player at any position from any year 
Round                  | The NFL draft round the player was selected Note:(1-7 being round, 8 being undrafted free agent)

```{r, echo = FALSE, warning = FALSE}
# Create a bar plot showing the number of players per year

nfl.complete = read.csv("career_combineNoNA.csv")
ggplot(nfl.complete, aes(x = as.factor(Year))) + 
  geom_bar(fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Number of NFL Quarterbacks per Draft Year(2000-2016)",
       x = "Year",
       y = "Number of Quarterbacks") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
```{r, echo = FALSE, warning = FALSE}
# Create a bar plot showing the number of players drafted in each round
ggplot(nfl.complete, aes(x = as.factor(Round))) + 
  geom_bar(fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of NFL Quarterbacks by Draft Round (2009-2016)",
       x = "Draft Round",
       y = "Number of Quarterbacks") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))
```
```{r, echo = FALSE, warning = FALSE}
# Create a single boxplot showing the overall distribution of experience
ggplot(nfl.complete, aes(y = experience)) + 
  geom_boxplot(fill = "lightblue", color = "black") +
  theme_minimal() +
  labs(title = "Overall Distribution of NFL Quarterbacks' Years Played",
       x = "",
       y = "Experience (Years)")
```
## Statistical Methods

=======
* Linear Discriminant Analysis (LDA)
  + Which is a statistical method for finding a linear combination of variables with the goal to classify new data into these categories. This technique is based on dimensionality reduction similar to principal component analysis. LDA allows us to make accurate predictions into multiple classification fields. In our case we applied LDA to determine a players draft round based on various parameters. 
  
* Principal Component Analysis (PCA)
  + This is a statistical technique used to reduce the number of variables by transforming the original variables into principal components. PCA allows us to reduce the dimensionality of a large data set. Note:Each subsequent component contains less variation within the data. In our case we applied PCA to reduce the dimensionality to discover variable relationships and importance. 
  
* Factor Analysis 
  + 
  
  
One of the parameters we aimed to be able to accurately predict is whether or not a given quarterback that is entering the draft will be drafted in the early rounds or not. The early rounds of the draft being the first and second as these are where you find most likely starting quarterbacks, and the later rounds being the remaining rounds being the third through seventh. To do this, the combine data for NFL quarterbacks was used to create an Linear Discriminant Analysis to predict whether the quarterbacks would be an early or later round draft picks. The variables used to predict the round drafted are Ht, Wt, Forty, Vertical, BroadJump, Cone, Shuttle, Year, and grade. The total data was split into a training and a testing set. The training set was used to create the LDA and the testing set was used to get estimated measurements for accuracy, specificity, and sensitivity. The training set was a random sample of about three fourths of the total data, and the testing was the remaining fourth. 

```{r}
# LDA analysis with combine data to predict if they are an early or late round draft pick

# Maybe modify to test whether they are a early round draft pick or most likely a starter for the team compared to a later pick most likely not a starter
set.seed(920981982)
combine.data <- read.csv("career_combineNoNA.csv")

espn.qb <- read.csv("espn.qb.csv")
espn.qb$qb <- espn.qb$player_name
total.combined <- combine.data %>% left_join(espn.qb, by = "qb")
combine.data <- total.combined %>% select(c(1:22, grade))
imp.1 <- mice(combine.data)
combine.data <- complete(imp.1)
# combine.data$grade[which(is.na(combine.data$grade))] <- mean(na.omit(combine.data$grade))

# Setting it to early or late round draft picks, Round 1 and 2 are early, 3 and beyond considered late
combine.data$Round <- ifelse(combine.data$Round==1 | combine.data$Round==2, "Early", "Late")
n <- sample(nrow(combine.data), size=75)
train.df <- combine.data[n,]
test.df <- combine.data[-n,]



lda.2 <- lda(Round~Ht+Wt+Forty+Vertical+BroadJump+Cone+Shuttle+Year+grade, data=combine.data)

pred.2 <- predict(lda.2, test.df)

correct.2 <- pred.2$class == test.df$Round

roc.2 <- roc(test.df$Round, as.numeric(pred.2$class))
roc.2$auc ###About 86.25% accuracy

roc.2$specificities ### Specificity about 87.5%
roc.2$sensitivities ### Sensitivity about 85%

# Predict quarterback rankings with LDA
s.darnold <- data.frame(Ht = 75, Wt = 221, Forty = 4.85, Vertical = 26.5, BroadJump = 105, Cone = 6.96, Shuttle = 4.4, Year = 2018, grade = 94)

predict(lda.2, s.darnold)$class



```
After calculating the LDA model and predicting the draft outcome for the testing set, we got an accuracy of about 86.25%, which is very good. The sensitivity, or the ability for the LDA to detect true positives, is about 87.5%. And finally the specificity, or the ability for the LDA to detect true negatives, is about 85%. These measures are what we hoped to see and give us reason to believe this model can be a good and accurate predictor of when during the draft, early or late, that an upcoming prospect will land based on the variables provided. To use a past example that isn't part of our data but does have all the data, we collected the variable measurements for Sam Darnold, an NFL quarter that was drafted in the Early rounds. Sam Darnold has the following data: Ht = 75, Wt = 221, Forty = 4.85, Vertical = 26.5, BroadJump = 105, Cone = 6.96, Shuttle = 4.4, Year = 2018, grade = 94. When we predicted when he would be drafted using our LDA it determined he would be taken in the early rounds. This was the correct prediction. 




```{r}
# Princomp
noncat2 = read.csv("career_combineNoNA_avgs.csv") %>%
  select(-1)
footballPC = princomp(noncat2[,2:21], cor = T)

#summary(footballPC, loadings = T) #.61 with 2 comps

autoplot(footballPC, loadings=T, loadings.label=T, data = noncat2)

# Dimension Rediction
#cortest.bartlett(cor(noncat2[,2:21]), n=103) #0 pval
#KMO(cor(noncat2[,2:21])) #Overall .82

fa.out = principal(noncat2[,2:21],nfactors=2,rotate="varimax")
print.psych(fa.out,cut=.5,sort=TRUE) #No cross-loading!

# fa.out = principal(noncat2[,2:21],nfactors=2,rotate="promax")
# print.psych(fa.out,cut=.5,sort=TRUE) #Both return same groups

# 2 distinct groupings, game stats and physical stats


#Group 1 pca (game stats)
  group1 = noncat2 %>%
    select(att, yds, cmp, game_points, td, sack, loss, int, AV, Round)
  
  
  g1PC = princomp(group1, cor = T)
  #summary(g1PC, loadings = T) #.92 with 2 comps
  
  autoplot(g1PC, loadings=T, loadings.label=T, data = group1)


#Group 2 pca (physical attributes)
  group2 = noncat2 %>%
    select(Forty, Vertical, BroadJump, Shuttle, Cone)
  
  
  g2PC = princomp(group2, cor = T)
  #summary(g2PC, loadings = T) #.81 with 2 comps
  
  autoplot(g2PC, loadings=T, loadings.label=T, data = group2) 
```

```{r}
season <- read.csv("qbStats.csv")
season.mlm <- lm(cbind(yds, td) ~ att+cmp+ypa+int+sack+loss+experience, data=season)
summary(season.mlm)

# Anova for the MLM shows that loss can be dropped from the model
Anova(season.mlm)

# Large p value from anova table with updated mlm shows that the 'loss' variable can be dropped because the model fits as well
season.mlm2 <- update(season.mlm, .~.-loss)
anova(season.mlm, season.mlm2)

# Test statistics from the linear hypothesis test show that the 'loss' variable can be dropped
lh.out <- linearHypothesis(season.mlm, hypothesis.matrix = c("loss = 0"))
lh.out
season.mlm2
qqnorm(season.mlm2$residuals[,1]); qqline(season.mlm2$residuals[,1])
qqnorm(season.mlm2$residuals[,2]); qqline(season.mlm2$residuals[,2])
```

